Hydrodynamics and rheology of adhesive vesicle doublets

### 2019-01-14 vesicle_code/docs/adhesion/adhesion.tex
some changes on e-flow and adhesion force

% ****** Start of file apssamp.tex ******
%
%   This file is part of the APS files in the REVTeX 4.1 distribution.
%   Version 4.1r of REVTeX, August 2010
%
%   Copyright (c) 2009, 2010 The American Physical Society.
%
%   See the REVTeX 4 README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
%
% See the REVTeX 4 README file
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex apssamp.tex
%  2)  bibtex apssamp
%  3)  latex apssamp.tex
%  4)  latex apssamp.tex
%
\documentclass[aps,prl,showpacs,amsmath,amssymb]{revtex4-1}

@@

\author{Bryan Quaife}
 \email{bquaife@fsu.edu}
 \altaffiliation[]{Department of Scientific Computing \\ Florida State University}
\author{Shravan Veerapaneni}%
 \email{shravan@umich.edu}
\altaffiliation{%
 Department of Mathematics \\
University of Michigan, Ann Arbor
}%
\author{Yuan-Nan Young}%
 \email{yyoung@njit.edu}
\altaffiliation{%
 Department of Mathematical Sciences \\
 New Jersey Institute of Technology
}%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Adhesion Model}
The adhesion model of Sukumaran and Seifert~\cite{suk-sei2001} considers adhesion
between a single three-dimensional vesicle and a solid wall at $z=0$.
\todo[inline]{Need a cartoon picture in here.  Some kind of version of
the one I had for my SIAM Life Sciences Talk}
Their model uses the adhesion potential
\begin{align*}
  \phi(z) = \sigma\left(\frac{\delta}{z}\right)^2 \left[ 
    \left(\frac{\delta}{z}\right)^2 - 2 \right],
\end{align*}
where $z$ is the distance between the solid wall and the vesicle,
$\sigma$ is the Hamaker constant, and $\delta$ is the adhesion length
scale.  We generalize this model to an unbounded suspension of vesicles.
We assume that the adhesion force applies between all pairs of points on
different vesicles, so we are interested in Hamaker constants that are
appropriate for lipid-lipid interactions. In
Appendix~\ref{sec:appendixA}, the adhesive force at a point $\xx$ on
vesicle $j$ is shown to be
\begin{align}
  \AA\xx:=-4 \sigma \delta^2\sum_{\substack{k=1 \\ k \neq j}}^M 
  \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^6} 
  \left(\delta^2 - \|\xx - \yy\|^2 \right) ds_\yy.
  \label{eqn:adhesionForce}
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Adhesion of two vesicles in an extensional flow} 
\label{sec:eflow} 
We consider two vesicles suspended in the extensional flow $\uu =
\chi(-x,y)$, where $\chi$ is the extensional rate.  The vesicles are
initially placed symmetrically on the $x$ axis in a quiescent flow
($\chi = 0$).  Once they reach steady-state doublet, the extensional
background flow is started.  We use this example to characterize the
required extensional rate to overcome the adhesive forces of the
doublet.  In particular, we investigate the effect of the reduced area.
For all the simulations, the length of the vesicles are fixed, while the
area is changed to achieve the desired reduced area.

@@

\begin{appendices}
\section{Adhesion Force}
\label{sec:appendixA}
Consider a suspension of two vesicles $\gamma_1$ and $\gamma_2$
parameterized as $\xx_1(s)$ and $\xx_2(s)$, respectively, with $s \in
[0,1]$.  Here $s$ is the arclength, and we have assumed, without loss of
generality, that both vesicles have length one.  We use the van der
Waals potential
\begin{align*}
  \phi(z) = \sigma\left(\frac{\delta}{z}\right)^2 \left[ 
    \left(\frac{\delta}{z}\right)^2 - 2 \right],
\end{align*}
where $z$ is the distance between two points.  Then, we define the total
adhesive energy on $\gamma_1$ to be
\begin{align*}
  U_1 = \int_{\gamma_1} \int_{\gamma_2} \phi(\|\xx_1 - \xx_2\|) 
    ds_{\xx_2} ds_{\xx_1}
\end{align*}
Perturbing $\xx_1$ to $\tilde{\xx}_1 = \xx_1 +  \delta \xx_1$, results
in a new vesicle $\tilde{\gamma}_1$, and the perturbed adhesive energy
is
\begin{align*}
  \widetilde{U}_1 = \int_{\tilde{\gamma}_1} \int_{\gamma_2}
  \phi(\|\tilde{\xx}_1 - \xx_2\|) ds_{\xx_2} ds_{\tilde{\xx}_1},
\end{align*}
and the change in the energy is
\begin{align*}
  \delta U_1 = \int_{\tilde{\gamma}_1} \int_{\gamma_2}
  \phi(\|\tilde{\xx}_1 - \xx_2\|) ds_{\xx_2} ds_{\tilde{\xx}_1} - 
  \int_{\gamma_1} \int_{\gamma_2} \phi(\|\xx_1 - \xx_2\|) 
  ds_{\xx_2} ds_{\xx_1}.
\end{align*}

We now decompose the perturbation into normal and tangential components
as $\delta \xx_1 = \epsilon \yy(s) = \epsilon(u\tt + v\nn)$. The
perturbed arclength term, to leading order, is
\begin{align*}
  \|\tilde{\xx}'_1\| \approx 1 + \epsilon(u_s + \nu\kappa),
\end{align*}
where $\kappa$ is the curvature.  To leading order, inextensible
perturbations satisfy $u_s + \kappa v = 0$, so the arclength term of
$\gamma_1$ and $\tilde{\gamma}_1$ are identical to leading order.
Therefore,
\begin{align*}
  \delta U_1 &= \int_{\gamma_1} \int_{\gamma_2} \left(
  \phi(\|\xx_1 + \epsilon \yy  - \xx_2\|) - \phi(\|\xx_1 - \xx_2\|)
  \right) ds_{\xx_2} ds_{\xx_1} \\
  &\approx \epsilon \int_{\gamma_1} \int_{\gamma_2}
  \nabla \phi (\|\xx_1 - \xx_2\|) \cdot \yy ds_{\xx_2} ds_{\xx_1}
\end{align*}
and the adhesive force applied by vesicle 2 on vesicle 1 is
\begin{align*}
  \int_{\gamma_2}\nabla \phi(\|\xx_1 - \xx_2\|)ds_{\xx_2} = 
  -4 \sigma \delta^2 \int_{\gamma_2}
  \frac{\xx_1 - \xx_2}{\|\xx_1 - \xx_2\|^6} 
  \left(\delta^2 - \|\xx_1 - \xx_2\|^2 \right) ds_{\xx_2}.
\end{align*}
A similar expression holds for the adhesive force applied by vesicle 1
on vesicle 2, and equation~\eqref{eqn:adhesionForce} gives the adhesive
force for a suspension of $M$ vesicles.

>>>>

% ****** Start of file apssamp.tex ******
%
%   This file is part of the APS files in the REVTeX 4.1 distribution.
%   Version 4.1r of REVTeX, August 2010
%
%   Copyright (c) 2009, 2010 The American Physical Society.
%
%   See the REVTeX 4 README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
%
% See the REVTeX 4 README file
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex apssamp.tex
%  2)  bibtex apssamp
%  3)  latex apssamp.tex
%  4)  latex apssamp.tex
%
\documentclass[aps,prl,showpacs,amsmath,amssymb,superscriptaddress]{revtex4-1}

@@

\author{Bryan Quaife}
\affiliation{Department of Scientific Computing, Florida State University, Tallahassee, Florida, USA}
% \email{bquaife@fsu.edu}
% \altaffiliation[]{Department of Scientific Computing \\ Florida State University}
\author{Shravan Veerapaneni}%
% \email{shravan@umich.edu}
\affiliation{Department of Mathematics, University of Michigan, Ann Arbor, Michigan, USA}%
\author{Y.-N. Young}%
 \email[Corresponding author: ]{yyoung@njit.edu}
\affiliation{Department of Mathematical Sciences, Newark, NJ, USA}%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Adhesion Model}
The adhesion model of Sukumaran and Seifert~\cite{suk-sei2001} considers adhesion
between a single three-dimensional vesicle and a solid wall at $z=0$.
\todo[inline]{Need a cartoon picture in here.  Some kind of version of
the one I had for my SIAM Life Sciences Talk}
The adhesion potential in their model is of the functional form in equation~\ref{eq:adhesion_potential} with $(m,n) = (4,2)$:
\begin{align}
\label{eq:adhesion_potential}
  \phi(z) = \sigma \left[ 
    \left(\frac{\delta}{z}\right)^m - \frac{m}{n} \left(\frac{\delta}{z}\right)^n \right],
\end{align}
where $z$ is the distance between the solid wall and the vesicle,
$\sigma$ is the Hamaker constant, and $\delta$ is the adhesion length
scale.  We generalize this model to an unbounded suspension of vesicles.
We assume that the adhesion force applies between all pairs of points on
different vesicles, so we are interested in Hamaker constants that are
appropriate for lipid-lipid interactions. In
Appendix~\ref{sec:appendixA}, the adhesive force at a point $\xx$ on
vesicle $j$ is shown to be
\begin{align}
  \AA\xx:=-\sigma m \delta^{n+1}\sum_{\substack{k=1 \\ k \neq j}}^M 
  \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^{m+1}} 
  \left(\delta^{m-n} - \|\xx - \yy\|^{m-n} \right) ds_\yy.
  \label{eqn:adhesionForce}
\end{align}

%\begin{align}
%  \AA\xx:=-4 \sigma \delta^2\sum_{\substack{k=1 \\ k \neq j}}^M 
%  \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^6} 
%  \left(\delta^2 - \|\xx - \yy\|^2 \right) ds_\yy.
%  \label{eqn:adhesionForce}
%\end{align}

The exponents $(m,n)$ depend on the geometry of the two objects under adhesion \cite{Book_IntermolecularSurfaceForces}: $(m,n)=(4,2)$ corresponds to two
flat, planar surfaces interacting with each under a long-range attraction and a short-range repulsion. On the other hand, $(m,n) = (12,6)$ corresponds to the L-J potential between
two molecules. The adhesion potential in  the integral of equation~\ref{eqn:adhesionForce} is between two small patches of lipid bilayer membrane. Thus we argue that
a reasonable choice for $(m,n)$ is between $(m,n) = (4,2)$ and $(m,n) = (12,6)$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Adhesion of two vesicles in an extensional flow} 
\label{sec:eflow} 
The hydrodynamics of a single vesicle in an 
extensional flow has revealed novel nonlinear vesicle dynamics not found for a viscous drop \cite{KantslerSegreSteinberg2008_PRL,ZhaoShaqfeh2013_JFM,Narsimhan2014_JFM,DahlNarsimhanGouveia2016_SoftMatt}.
The stagnation point (zero-velocity point) in an extensional flow is a saddle point where two laminar streams converge along an axis and then diverge in a
different direction. 
With an active control algorithm to fix the position of the stagnation point by adjusting the streaming flow strength with a feedback loop \cite{Johnson-Chavarria2011_EMJ},
a particle can be trapped at the stagnation point for long time scales to facilitate
image acquisition or other detailed measures such as PIV of flow inside/around the particle.
To explore the application of such fluid trap to measure the adhesion strength between two bound vesicle membranes, we propose the following thought experiment.

We begin from two vesicles forming a doublet under an adhesion potential in a quiescent flow as in \S~\ref{sec:qflow}. 
Once a doublet forms and reaches an equilibrium configuration, we turn on 
the fluid trap with the stagnation point placed at the center of the vesicle doublet.
For simplification we assume that each vesicle in the doublet has the same reduced area, membrane area, and bending stiffness.
At low extensional rate, we expect the vesicle doublet to stay bound at the fluid trap stagnation point.
On the other hand, the vesicle doublet may become unstable and eventually separate at a high extensional rate.
Thus we expect there to exist a critical extensional rate above which the vesicle doublet cannot stay bound under a given adhesion potential.
The dependence of the critical extensional rate on the adhesion potential and the mechanical vesicle properties thus provides a means to probe the physics of membrane adhesion.

We consider two vesicles suspended in the extensional flow $\uu =
\chi(-x,y)$, where $\chi$ is the extensional rate.  The vesicles are
initially placed symmetrically on the $x$ axis in a quiescent flow
($\chi = 0$).  Once they reach steady-state doublet, the extensional
background flow is started.  We use this example to characterize the
required extensional rate to overcome the adhesive forces of the
doublet.  In particular, we investigate the effect of the reduced area.
For all the simulations, the length of the vesicles are fixed, while the
area is changed to achieve the desired reduced area.

@@

\begin{appendices}
\section{Adhesion Force}
\label{sec:appendixA}
Consider a suspension of two vesicles $\gamma_1$ and $\gamma_2$
parameterized as $\xx_1(s)$ and $\xx_2(s)$, respectively, with $s \in
[0,1]$.  Here $s$ is the arclength, and we have assumed, without loss of
generality, that both vesicles have length one.  We use the van der
Waals potential
\begin{align*}
  \phi(z) = \sigma \left[ 
    \left(\frac{\delta}{z}\right)^m - \frac{m}{n}\left(\frac{\delta}{z}\right)^n \right],
\end{align*}
where $z$ is the distance between two points.  Then, we define the total
adhesive energy on $\gamma_1$ to be
\begin{align*}
  U_1 = \int_{\gamma_1} \int_{\gamma_2} \phi(\|\xx_1 - \xx_2\|) 
    ds_{\xx_2} ds_{\xx_1}
\end{align*}
Perturbing $\xx_1$ to $\tilde{\xx}_1 = \xx_1 +  \delta \xx_1$, results
in a new vesicle $\tilde{\gamma}_1$, and the perturbed adhesive energy is
\begin{align*}
  \widetilde{U}_1 = \int_{\tilde{\gamma}_1} \int_{\gamma_2}
  \phi(\|\tilde{\xx}_1 - \xx_2\|) ds_{\xx_2} ds_{\tilde{\xx}_1},
\end{align*}
and the change in the energy is
\begin{align*}
  \delta U_1 = \int_{\tilde{\gamma}_1} \int_{\gamma_2}
  \phi(\|\tilde{\xx}_1 - \xx_2\|) ds_{\xx_2} ds_{\tilde{\xx}_1} - 
  \int_{\gamma_1} \int_{\gamma_2} \phi(\|\xx_1 - \xx_2\|) 
  ds_{\xx_2} ds_{\xx_1}.
\end{align*}

We now decompose the perturbation into normal and tangential components
as $\delta \xx_1 = \epsilon \yy(s) = \epsilon(u\tt + v\nn)$. The
perturbed arclength term, to leading order, is
\begin{align*}
  \|\tilde{\xx}'_1\| \approx 1 + \epsilon(u_s + \nu\kappa),
\end{align*}
where $\kappa$ is the curvature.  To leading order, inextensible
perturbations satisfy $u_s + \kappa v = 0$, so the arclength term of
$\gamma_1$ and $\tilde{\gamma}_1$ are identical to leading order.
Therefore,
\begin{align*}
  \delta U_1 &= \int_{\gamma_1} \int_{\gamma_2} \left(
  \phi(\|\xx_1 + \epsilon \yy  - \xx_2\|) - \phi(\|\xx_1 - \xx_2\|)
  \right) ds_{\xx_2} ds_{\xx_1} \\
  &\approx \epsilon \int_{\gamma_1} \int_{\gamma_2}
  \nabla \phi (\|\xx_1 - \xx_2\|) \cdot \yy ds_{\xx_2} ds_{\xx_1}
\end{align*}
and the adhesive force applied by vesicle 2 on vesicle 1 is
\begin{align}
 \int_{\gamma_2}\nabla \phi(\|\xx_1 - \xx_2\|)ds_{\xx_2} = -m \sigma\delta^{n+1}
   \int_{\gamma_k} \frac{\xx - \yy}{\|\xx - \yy\|^{m+1}} 
  \left(\delta^{m-n} - \|\xx - \yy\|^{m-n} \right) ds_\yy.
  \end{align}
When $(m,n) = (4,2)$, the above express becomes
\begin{align*}
  \int_{\gamma_2}\nabla \phi(\|\xx_1 - \xx_2\|)ds_{\xx_2} = 
  -4 \sigma \delta^2 \int_{\gamma_2}
  \frac{\xx_1 - \xx_2}{\|\xx_1 - \xx_2\|^6} 
  \left(\delta^2 - \|\xx_1 - \xx_2\|^2 \right) ds_{\xx_2}.
\end{align*}
A similar expression holds for the adhesive force applied by vesicle 1
on vesicle 2, and equation~\eqref{eqn:adhesionForce} gives the adhesive
force for a suspension of $M$ vesicles.
